Is high social class always beneficial for survival? A study of adults and

### 2017-09-22 article2.Rmd
trivia

---
title: Life course and long-term perspectives of social inequality in mortality among
  elderly and adults in Northern Sweden      1801–2013
author: "S&ouml;ren Edvinsson and G&ouml;ran Brostr&ouml;m"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    keep_tex: true
    citation_package: natbib
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
    toc_float: yes
bibliography: mybib.bib
documentclass: article
fontsize: 12pt
citation_package: natbib
biblio-style: apalike
---

@@

A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality. In the present study, we challenge common notions of this by taking a very long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences. The arguments for our statements is based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the period 1801--2013 for Skellefteå and 1901--2013 for Umeå. The main issue is then analysed according to gender and age-group (working age vs retired), testing if the same patterns prevails in old age -- the retired population -- versus the population in working age. We analyse both total and mortality and mortality from the major causes of death (cardiovascular diseases and cancers). Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society. The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.

@@

One of the central aspects of survival is social class and access to economic
and other resources. Even in present-day welfare societies, social position is a
strong determinant when it comes to health and mortality and the impact even
seems to be increasing [@kuetal04, @macketal16, @frlu07, @broba12, @strand10]. It has been suggested that
"... social conditions have been, are and will continue to be irreducible
determinants of health outcomes and therefore deserves their appellation as
'fundamental causes' of disease and death" [@liph95]. The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in this theory [@liph98]. For a long time, the general view has been that socio-economic health and mortality inequalities were high in
historical societies, probably higher than in modern societies. This is a
reasonable assumption since these societies were in most cases characterised by
very large socioeconomic differences. Knowing that access to
resources provides advantages in all aspects of life, the health advantage of higher
classes ought to be obvious. @ant67 suggests that social
inequality in mortality has passed through different historical phases. According
to him, differences were comparatively small during the
pre-transitional phase. This period was characterised by space being a strong determinant for the spread of disease. Differences then increased during the transitional phase when mortality
declined and wealthy groups used their economic and human resources to gain better health. Finally,
mortality differences decreased again resulting in marginal levels of inequality in modern low-mortality societies when instead
health-related behaviour became the decisive determinant for health and survival.
@omran82 comes to a similar conclusion in his theory of the epidemiologic transition. He states in the third proposition of the theory that even if the class differentials in mortality were
maintained during the transition, the decline set in earlier and was faster
among privileged groups.

Recent studies, investigating social inequality in health and mortality
with micro-data, have however questioned how general the assumed
pattern is [@bevp11]. Solid empirical evidence about the process is however lacking and studies
focusing on the issue are still few. There is a need for additional reliable
studies from different geographical and historical settings in order to 
better understand the role of socio-economic conditions for health and survival over time.

What about life course aspects of the impact of social position for health and mortality? There are
identified three main hypotheses concerning social health inequalities in adulthood versus old age [@hof08]. Either differences converge in old age (status levelling), the differences are constant (status maintenance), or they diverge (cumulative advantage). Diminishing
differences may be a consequence of the circumstance that biological factors play a stronger role than social factors later in life. The status maintenance hypothesis basically assume 
continuity in the determinants for social health inequalities from adulthood to old age.
The third is the cumulative (dis)advantage hypothesis [@dan03], implying that advantages and disadvantages tend to persist and ackumulate during life, constantly rewarding some while disfavouring others. This would lead to larger differences in old age.

Another aspect of the development of social inequality in mortality,
concerns its relation to economic inequality. @wipi09 argue that
income inequality has an independent effect on mortality, separate from the
direct effect of actual access to economic resources. They argue
that unequal societies (basically countries) perform less well when it comes to health (as well as other social conditions) than equal ones in the present-day economically developed world. This has initiated a vital scholarly debate and the topic has been extensively studied [@suka04,
@wado00]. During the last decades, the association between trends of inequalities in mortality and income respectively are weak or non-existant according to @hoetal16. How the association looked like in previous periods is unknown. This is of particular interest since the levels of economic
inequality were very different from those of the recent decades. Even if not a necessary implication, it is reasonable to assume that poorer groups were those most disadvantaged of large inequalities.

@@

```{r mapp,fig.cap = "The Skellefteå and Umeå regions in Sweden.", out.width="3in", out.height = "4.1in", fig.align="center"}
knitr::include_graphics("myfigs/sweden.png")
```
The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the
county of Västerbotten in the north of Sweden along the coast of Gulf
of Bothnia. This was a remote part of the country where communication with
the rest of Sweden was difficult until the late 19th century. The economy
was dominated by agriculture, making it vulnerable to harvest failures, and
several severe famines frequently occurred in the regions during the 1800s, for example after the harvest failure in 1867 [@segb14]. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. The first ships during the year came to this area in May or even as
late as June in some cases [@fah56]. Towards the end of the 19th
century, the Swedish railway system reached this part of Sweden. Thereby
contacts with the rest of Sweden were facilitated, improving the economy and making it possible to mitigate the effects of harvest failures. During the 19th century, our regions became
increasingly integrated in the same epidemiological pattern as the rest of
Sweden.

In our dataset, the Skellefteå region before 1950 consists of a selection
of parishes surrounding the town of Skellefteå, founded in 1845 but with a
very small population during the 19th century. The data from the period
after 1975 cover the Skellefteå, Norsjö and Malå municipalities, the same area as
for the earlier period but with the addition of two more parishes. The
majority of the 19th century population lived in rural villages and
hamlets, getting their livelihood from agricultural production. During the
20th century, industrialisation took place. This also led to a population increase both in the town
and in the rural parts, resulting in a much more diversified economy.

@@

The Umeå region in the dataset consists of Umeå urban and rural parish 1901--1950, and from 1976 onwards of Umeå municipality when another three
parishes are included. This region has a somewhat different character from
that of Skellefteå. Umeå town had a small population but was substantially
larger than Skellefteå town during almost the whole studied period. It was
the administrative centre in the county of Västerbotten. Schools and
military regiments were placed here and the economy was more diversified
compared to Skellefteå. Agriculture dominated the rural part, but there
were also some foundries as well as industries for example in forestry and
small-scale production. The population size as defined in our data sets (all ages) was
19138 on January 1, 1901, 33393 on December 31, 1950 and 103970 when
the 20th century ended.

@@

The other large data-set is extracted from the Linnaeus database
[@malm10]. Linnaeus is based on different linked national population
registers from 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
University. The study period from 1976 to 2000 are constructed from
censuses every fifth year 1976--1990, with additional information on deaths
from National Board of Health and Welfare (we have left out the period 1960--1975 due to more missing data in some variables, for example social class, and additional work required for defining the exact geographical areas for analysis). For the period 2002--2013 we use
the information from the yearly population registers from Statistics Sweden
(LISA) together with information on mortality from National Board of Health and
Welfare.

@@

```{r lexis2,fig.cap="The sampling frame.", fig.height = 5}
plot(c(1801, 1801, 1951, 1951, 1801), c(40, 90, 90, 40, 40),
     type = "l", xlim = c(1800, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
axis(1, at = c(1801, 1851, 1901, 1951, 1976, 1991, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1976, 1976, 2014, 2014, 1976), c(40, 90, 90, 40, 40),
      col = "blue", lwd = 2)
#
pn.exp <- round(sum(skel$exposure) + sum(skume$exposure))
pn.d <-  as.integer(sum(skel$event) + sum(skume$event))
#
text(1875, 70, "PopLink", cex = 2, col = "red")
##text(1899, 60, paste(pn.rec, "records"))
text(1875, 55, paste(pn.exp, "p-years"))
text(1874.5, 50, paste(pn.d, "deaths"))
#
n.exp <- round(sum(foli$exposure))
n.d <-  as.integer(sum(foli$event))

text(1995, 70, "Linnaeus", cex = 1.5, col = "blue")
##text(1984, 60, paste(n.rec, "records"))
text(1995.2, 55, paste(n.exp, "p-years"), cex = 0.8)
text(1995.2, 50, paste(n.d, "deaths"))
```

@@

Differences in available information in the two datasets as well as within
the Linnaeus data makes it necessary to apply different approaches when it
comes to identification of presence periods. The Poplink data provides us with exact dates or at least year of start and exit of the presence of all individuals in our data, allowing us
to have full and continuous control of the \emph{de jure} population. This is
however not the case with the Linnaeus database. For the period 1976--2000
we use presence in the censuses 1976--1990 and information on deaths from the National
Board of Health and Welfare to identify the observation periods in our
analyses. There is however no full information on what happened in-between
the censuses, for example the exact date of exit for persons having moved
between two censuses. We have solved this by considering everyone as
resident in our regions during the complete five-year period between
censuses if the person is missing in the later census (the last period covering ten years). This marginally overestimates presence, but should not bias the results when it
comes to socioeconomic differences. We furthermore miss those that both
moved in and moved out or vice versa between censuses. Exit dates are exact
if they refer to death. In these cases we consider them as constantly
present from the last census to the death data even if the death took place
in another part of Sweden.

@@

In this paper we analyse mortality differences according to social class, and based on a modified form of the classification scheme Hisclass [@lema11]. The only
type of socio-economic information available both in the Linnaeus and Poplink database
is occupations. Thus, analysing according to social class is apart from
being a scientific choice, also a practical one. Using occupational data is
the only and obvious alternative for making a longitudinal study of
socio-economic mortality differences with consistent classification
schemes. Social position is treated as a time-dependent variable during working age but defined from the last occupation for the elderly and retired population, i.e. from age 65 until death or last observation.

The availability to information on occupations varies over time and between
sources. Occupations in the version of Poplink we are using are coded according to the original DDB coding system that then has been adapted to Hisclass. For the
Linnaeus data, we use a script recoding HISCO into Hisclass. The censuses include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
Yrkesklassificering). We consider the observed social class in a census to be valid for the full period until the next census, corresponding to our definition of presence periods. For the period 1991--2000 the social class in the census 1990 is considered to be kept until 2001.

@@

Female labour is much better covered in the Linnaeus database. A
complication is however that it is  difficult to define households in the
same way as in Poplink. Extra-marital cohabitation became common and female
labour force participation developed as the norm in Sweden during this
period. The occupation of the husband as signifier of social class became,
if not obsolete, at least less relevant. All included persons are signified
by their individual occupation for this period. Still, the results for the
different periods are not completely comparable and makes the
interpretation of the development for women uncertain for some periods. Notice however that there probably is a strong social homogeneity between partners in a marriage, indicating that the results would not differ in any decisive way by using occupations on individual level instead of on household level for defining social class from the Linneaus data [@bjork92].

@@

Figures \@ref(fig:hiscskumm) and \@ref(fig:hiscskumw) show the distribution
(per cent) of exposure time according to class for men and women
respectively. The category of workers is fairly stable over time. The
highest class has increased substantially, although from a very low level
in the 19th century. The increase of academics in connection to the
establishment of Umeå University explains much of this increase. What is
not apparent from Figure \@ref(fig:hiscskumm) is the transfer within the middle
group from a majority of farmers during the 19th century to lower white
collar from the middle of the 20th century. Furthermore we observe a rather high percentage of missing social class in the early 19th century. For men the proportion is somewhat higher in the age group 65 and above, i.e. those that were retired, but still low enough to be used in our analyses (results not shown). Until 1950, there were however very few with missing occupation.

@@

The proportion with missing values among women is however a larger problem,
see Figure \@ref(fig:hiscskumw). Corresponding to what we found for men, the
proportion without social position is very low until the middle of the 20th
century, reflecting the way womens' social position is defined. Missing social class is however quite high during particularly the first two periods with data from the Linnaeus database -- from 1976. Since female social class for this period is defined
according to own occupation, the consequence of many women being outside
the workforce is that the proportion missing becomes very high. For the later periods,
the results are fairly complete.

@@

Regarding the distribution of marital status over time, see
Figure \@ref(fig:civstskum).

@@

In 19th and early 20th century Sweden (as in other countries), mortality
was much higher in urban environments [@hani94, @edv92]. The unsanitary and overcrowded
living conditions as well as the risks for importing infections made the
health risks higher in towns and cities. In the present paper, the variable **urban** controls
for this aspect by distinguishing between urban and rural residence for the periods after 1900, based
on parish or place of residence.

@@

We have made separate analyses for the different periods. The century
1801--1950 is divided into six periods of 25 years each. The 19th century
periods represent a mainly pre-industrial society dominated by agriculture
and with only basic welfare provisions. During the years 1901--1950,
industrialisation started and the urban environments increased in
population size. The modern Swedish welfare state began to develop. For the 
period 1976 onwards, there are separate analyses on decades except for the two 
last periods. Sweden had developed into a wealthy
welfare society with low income inequality,
particularly until the 1990s. A financial crisis struck Sweden in the early
1990s, which among other things increased unemployment and changed parts of
the welfare system towards being less generous. The start of the
last period (2008--2013) coincides with the international financial
crisis. During the last decades income inequality has increased considerably.

@@

# The effect of social class, 1801--1950, 1976--2013

@@

The models are fitted separately for sex and each period, from  1801 to
2013. We stratify on social class (**hisclass**) and include the covariates **urban** (TRUE or FALSE) and
**civst** (marital status). The cumulative
hazards at the end of the age period (65 if 40--64, 90 if 65--89) are the
main target in our analysis. These numbers are used in the process of
graphically illustrating the development over time.

@@

Notice that we have not analysed the 1950s due to the fact that we only
have data for some parishes at that time. The
changing economy have some consequences for the analysis, particularly when
it comes to the agricultural sector. While farmers were common occupations
up until the first half of the 20th century, these occupations has constantly
decreased in numbers thereafter. This means that the middle class has
a different character during the last periods.

The most striking result of the analysis is the different patterns for men
and women. The social pattern of mortality among women is quite much in
accordance with previous analyses from the present research group on other
regions in northern Sweden and for other periods [@edli11, @edbr12]. The
highest social class usually have comparatively low mortality, while
working class women have high, but almost constantly about the same level as the middle category. There is however no substantial change in the relative risks between classes. Only the group with unknown social class is distinguished by much higher mortality from the 1980s
onwards. This can however have to do with how we have defined social
position for women before and after 1950. After 1975 it is based
on their own occupation. Being without occupation probably include women that
are more frail when it comes to health or are vulnerable due to social characteristics.

A completely different pattern appears among men, being the most surprising finding in
our analyses. The groups we would expect to have the lowest mortality, if
we assume that access to vital resources and belonging to a high social class
determines survival, instead have by far the highest hazards of dying
during the nineteenth and much of the 20th century. Workers on the
other hand have comparatively low mortality until 1950, at the same
level as the middle class. The fact that we define social class as a time-dependent variable could otherwise lead to reverse causation -- those having bad health losing social position -- but instead we still find better survival among workers compared to the elite.

The relative positions change radically towards
the end the 20th century. From the 1980s, there is a clear advantage when
it comes to survival for higher managers and professionals. The relative
differences are however more compressed than for the period until 1950. The
expected social gradient in mortality has developed. This is also what we
usually find in analyses of social health inequalities in present-day
Sweden as well as in most other countries. In the regions we are studying,
this is however a quite recent phenomenon. Corresponding to what we find
for women, those without known occupation have much higher mortality than
the others from the late 1970s onwards. The assumption made for women of
unknown social class is probably even more relevant for men. They probably
represent groups in society that are particularly vulnerable.

@@

Concerning the social pattern, the pattern is less clear compared to what
we find for the younger age group in line with the status levelling hypothesis (see Figure \@ref(fig:cumswold)). For
women the results are mixed. We do not find the
same clear advantage for elite women during the period before 1950, but neither
do we find any large disadvantage for working class elderly women at that
time. In some periods, the social differences were small, particularly between the worker and the middle group, and the elite even had substantially higher mortality during the early 20th century. The
social differences continued to be small until the last decades of the
20th century. Only during the two last periods (after the year 2000) is it
possible to establish the expected social gradient in mortality.

@@

In this and the following section, the focus is on analyses of mortality from the two
dominant groups of death causes in the adult and elderly population, using the same models and presentations as for total mortality. We restrict the analyses to the period 1901 on wards. We start
with cardiovascular diseases. Many studies have shown that socio-economic
conditions are strongly associated with mortality in this disease
group. Belonging to working class, having less education and low income
increases mortality risks. We would therefore expect to find a clear
social pattern in the present analysis.

@@

@@

Figure \@ref(fig:41cumsw) shows mortality levels during different periods
from 1901 on-wards. Women from the highest social class had consistently
lower mortality from cardiovascular diseases compared to the others. We
also observe increasing social differences from the first half of the
century to the last decades. A new pattern has developed from 1980
onwards. Now working class women have substantially higher cardiovascular
mortality compared to others. We can however not exclude that this is an
artefact of different diagnostics for deaths in different groups. Working
class women may have had less precise diagnoses in the early 20th century.

For men, there is a different pattern being to a large
extent consistent with what we observed concerning all-cause
mortality. The elite group had much higher cardiovascular mortality up until
the middle of the 20th century, but instead have the lowest mortality from
1980 onwards. Working class men had a fairly low cardiovascular mortality,
but in the same way as among women, the levels were fairly similar to the
one of the middle category consisting of white collar, lower managers and farmers. In the last decades and similarly as for women, working class men have the highest
mortality from this cause of death group. A social
hierarchy in cardiovascular mortality developed both for men
and women.

@@

# A cohort approach, all causes of death

>>>>

---
title: Is high social class always beneficial for survival? A study of adults and
  elderly in Northern Sweden 1801--2013
author: "S&ouml;ren Edvinsson and G&ouml;ran Brostr&ouml;m"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::pdf_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
    toc_float: yes
bibliography: mybib.bib
documentclass: article
citation-package: natbib
biblio-style: apalike
---

@@

A major theme in demographic and epidemiological studies is the seemingly persistent effect of social class on mortality. In the present study, we challenge this notion by taking a very long-term perspective on the development of social inequalities in mortality in the adult and elderly population with a special focus on the relative differences. The arguments for our statements is based on an investigation of the Skellefteå and Umeå regions in the north of Sweden for the period 1801--2013 for Skellefteå and 1901--2013 for Umeå. The main issue is then analysed according to gender and age-group (working age vs retired) and major causes of death (cardiovascular diseases and cancers). Furthermore we put this in the context of how the inequality in mortality is associated with the development of economic inequality in society. The results are discussed in relation to our understanding of the mortality transition and the social determinants of health and mortality, as well as their implications on some of the most influential hypotheses and concepts in health research.

@@

One of the central aspects of survival is social class and access to economic
and other resources. Even in present-day welfare societies, social position is a
strong determinant when it comes to health and mortality and the impact even
seems to be increasing [@kuetal04, @macketal16, @frlu07, @broba12, @strand10]. It has been suggested that
"... social conditions have been, are and will continue to be irreducible
determinants of health outcomes and therefore deserves their appellation as
'fundamental causes' of disease and death" [@liph95]. The persistence of social inequality in mortality to the disadvantage of the lower classes is one of the main assumptions in this theory [@liph98]. For a long time, the general view has been that socio-economic health and mortality inequalities were high in
historical societies, probably higher than in modern societies. This is a
reasonable assumption since these societies were in most cases characterised by
very large socioeconomic differences. Knowing that access to
resources provides advantages in all aspects of life, the health advantage of higher
classes ought to be obvious. @ant67 suggests that social
inequality in mortality has passed through different historical phases. According
to him, differences were comparatively small during the
pre-transitional phase when space was a strong determinant for the spread of
disease. Differences then increased during the transitional phase when mortality
declined and wealthy groups used their economic and human resources to gain better health. Finally,
mortality differences decreased again resulting in marginal levels of inequality in modern low-mortality societies when instead
health-related behaviour became the decisive determinant for health and survival.
@omran82 comes to a similar conclusion in his theory of the epidemiologic transition. He states in the third proposition of the theory that even if the class differentials in mortality were
maintained during the transition, the decline set in earlier and was faster
among privileged groups.

Recent studies, investigating social inequality in health and mortality
with micro-data, have however questioned how general the assumed
pattern is [@bevp11]. Solid empirical evidence about the process is however lacking and studies
focusing on the issue are still few. There is a need for more reliable
studies from different geographical and historical settings in order to 
better understand the role of socio-economic conditions for health and survival over time.

Another aspect of the development of social inequality in mortality,
concerns its relation to economic inequality. @wipi09 argue that
income inequality has an independent effect on mortality, separate from the
direct effect of actual access to economic resources. They argue
that unequal societies (basically countries) perform less well when it comes to health (as well as other social conditions) than equal ones in the present-day economically developed world. This has initiated a vital scholarly debate and the topic has been extensively studied [@suka04,
@wado00]. During the last decades, the association between trends of inequalities in mortality and income respectively are weak or non-existant according to @hoetal16. How the association looked like in previous periods is unknown. This is of particular interest since the levels of economic
inequality were very different from those of the recent decades. Even if it
not a necessary implication, it is reasonable to assume that poorer groups were
those most disadvantaged of large inequalities.

@@

```{r mapp,fig.cap = "The Skellefteå and Umeå regions in Sweden.", out.width="3in"}
knitr::include_graphics("myfigs/sweden.png")
```
The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the
county of Västerbotten in the north of Sweden along the coast of Gulf
of Bothnia. This was a remote part of the country where communication with
the rest of Sweden was difficult until the late 19th century. The economy
was dominated by agriculture, making it vulnerable to harvest failures, and
several severe famines frequently occurred in the regions during the 1800s, for example after the harvest failure in 1867 [@segb14]. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. The first ships during the year came to this area in May or even as
late as June in some cases [@fah56]. Towards the end of the 19th
century, the Swedish railway system reached this part of Sweden. Thereby
contacts with the rest of Sweden were facilitated, improving the economy and making it possible to mitigate the effects of harvest failures. The inland was particularly
remote and isolated. During the 19th century, our regions became
increasingly integrated in the same epidemiological pattern as the rest of
Sweden.

In our dataset, the Skellefteå region before 1950 consists of a selection
of parishes surrounding the town of Skellefteå, founded in 1845 but with a
very small population during the 19th century. The data from the period
after 1960 cover the Skellefteå and Malå municipalities, the same area as
for the earlier period but with the addition of two more parishes. The
majority of the 19th century population lived in rural villages and
hamlets, getting their livelihood from agricultural production. During the
20th century, industrialisation took place. This also led to a population increase both in the town
and in the rural parts, resulting in a much more diversified economy.

@@

The Umeå region in the dataset consists of Umeå urban and rural parish 1901--1950, and from 1960 onwards of Umeå municipality when another three
parishes are included. This region has a somewhat different character from
that of Skellefteå. Umeå town had a small population but was substantially
larger than Skellefteå town during almost the whole studied period. It was
the administrative centre in the county of Västerbotten. Schools and
military regiments were placed here and the economy was more diversified
compared to Skellefteå. Agriculture dominated the rural part, but there
were also some foundries as well as industries for example in forestry and
small-scale production. The population size as defined in our data sets (all ages) was
19138 on January 1, 1901, 33393 on December 31, 1950 and 103970 when
the 20th century ended.

@@

The other large data-set is extracted from the Linnaeus database
[@malm10]. Linnaeus is based on different linked national population
registers from 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
University. The study period from 1960 to 2000 are constructed from
censuses every fifth year 1960--1990, with additional information on deaths
from National Board of Health and Welfare. For the period 2002--2013 we use
the information from the yearly population registers from Statistics Sweden
(LISA) together with information on mortality from National Board of Health and
Welfare.

@@

```{r lexis2,fig.cap="The sampling frame.", fig.height = 5}
plot(c(1801, 1801, 1951, 1951, 1801), c(40, 90, 90, 40, 40),
     type = "l", xlim = c(1800, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
axis(1, at = c(1801, 1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
      col = "blue", lwd = 2)
#
pn.exp <- round(sum(skel$exposure) + sum(skume$exposure))
pn.d <-  as.integer(sum(skel$event) + sum(skume$event))
#
text(1875, 70, "PopLink", cex = 2, col = "red")
##text(1899, 60, paste(pn.rec, "records"))
text(1875, 55, paste(pn.exp, "p-years"))
text(1874.5, 50, paste(pn.d, "deaths"))
#
n.exp <- round(sum(foli$exposure))
n.d <-  as.integer(sum(foli$event))

text(1985, 70, "Linnaeus", cex = 2, col = "blue")
##text(1984, 60, paste(n.rec, "records"))
text(1985.2, 55, paste(n.exp, "p-years"))
text(1985.2, 50, paste(n.d, "deaths"))
```

@@

Differences in available information in the two datasets as well as within
the Linnaeus data makes it necessary to apply different approaches when it
comes to identification of presence periods. The Poplink data provides us with exact dates or at least year of start and exit of the presence of all individuals in our data, allowing us
to have full and continuous control of the \emph{de jure} population. This is
however not the case with the Linnaeus database. For the period 1960--2000
we use presence in the censuses 1960--1990 and information on deaths from the National
Board of Health and Welfare to identify the observation periods in our
analyses. There is however no full information on what happened in-between
the censuses, for example the exact date of exit for persons having moved
between two censuses. We have solved this by considering everyone as
resident in our regions during the complete five-year period between
censuses if the person is missing in the later census (the last period covering ten years). This marginally overestimates presence, but should not bias the results when it
comes to socioeconomic differences. We furthermore miss those that both
moved in and moved out or vice versa between censuses. Exit dates are exact
if they refer to death. In these cases we consider them as constantly
present from the last census to the death data even if the death took place
in another part of Sweden.

@@

In this paper we analyse mortality differences according to social class, treated as a time-dependent variable, and based on a modified form of the classification scheme Hisclass [@lema11]. The only
type of socio-economic information available both in the Linnaeus and Poplink database
is occupations. Thus, analysing according to social class is apart from
being a scientific choice, also a practical one. Using occupational data is
the only and obvious alternative for making a longitudinal study of
socio-economic mortality differences with consistent classification
schemes. Social position is treated as a time-dependent variable.

The availability to information on occupations varies over time and between
sources. Occupations in the version of Poplink we are using are coded according to the original DDB coding system that then has been adapted to Hisclass. For the
Linnaeus data, we use a script recoding HISCO into Hisclass. The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk
Yrkesklassificering). We consider the observed social class in a census to be valid for the full period until the next census, corresponding to our definition of presence periods. For the period 1991--2000 the social class in the census 1990 is considered to be kept
until 2001. Furthermore, we consider the last social position before retirement
to be kept until death or last observation for those above 65.

@@

Female labour is much better covered in the Linnaeus database. A
complication is however that it is  difficult to define households in the
same way as in Poplink. Extra-marital cohabitation, connected to the Second
Demographic Transition [@van87], became common and female
labour force participation developed as the norm in Sweden during this
period. The occupation of the husband as signifier of social class became,
if not obsolete, at least less relevant. All included persons are signified
by their individual occupation for this period. Still, the results for the
different periods are not completely comparable and makes the
interpretation of the development for women uncertain for some periods. Notice however that there probably is a strong social homogeneity between partners in a marriage, indicating that the results would not differ in any decisive way by using occupations on individual level instead of on household level for defining social class from the Linneaus data [@bjork92].

@@

Figures \@ref(fig:hiscskumm) and \@ref(fig:hiscskumw) show the distribution
(per cent) of exposure time according to class for men and women
respectively. The category of workers is fairly stable over time. The
highest class has increased substantially, although from a very low level
in the 19th century. The increase of academics in connection to the
establishment of Umeå University explains much of this increase. What is
not apparent from Figure \@ref(fig:hiscskumm) is the transfer within the middle
group from a majority of farmers during the 19th century to lower white
collar from the middle of the 20th century. Furthermore we observe an
increase during the 1960s and partly during later decades as well in those
with undefined class. When first observed in the earliest included censuses,
occupation is often missing. For men the proportion is somewhat higher in
the age group 65 and above, i.e. those that were retired, but still low
enough to be used in our analyses (results not shown). Until 1950, there were however very few
with missing occupation. For men in working-active age, the proportion
missing does not cause any problems for our analyses.

@@

The proportion with missing values among women is however a larger problem,
see Figure \@ref(fig:hiscskumw). Corresponding to what we found for men, the
proportion without social position is very low until the middle of the 20th
century, reflecting the way womens' social position is defined. Missing social class is however a real problem during particularly the 1960s and 1970s. Around 80 \% lack information for the 1960s and 50 \% for the 1970s. Since social class for this period is defined
according to own occupation, the consequence of many women being outside
the workforce, the proportion missing becomes very high. We have chosen not
to report results for women for these two decades. For the later decades,
the results are fairly complete.

@@

Regarding the distribution of marital status over time, see
Figure \@ref(fig:civstskum). The figure indicates an under-registration of
those married and over-registration of unmarried during the
1960s.

@@

In 19th and early 20th century Sweden (as in other countries), mortality
was much higher in urban environments [@hani94, @edv92]. The unsanitary and overcrowded
living conditions as well as the risks for importing infections made the
health risks higher in towns and cities. In the present paper, we control
for this aspect by distinguishing between urban and rural residence for the periods after 1900, based
on parish or place of residence.

@@

We have made separate analyses for the different periods. The century
1801--1950 is divided into six periods of 25 years each. The 19th century
periods represent a mainly pre-industrial society dominated by agriculture
and with only basic welfare provisions. During the years 1901--1950,
industrialisation started and the urban environments increased in
population size. The modern Swedish welfare state began to develop. For the 
period 1960 onwards, there are separate analyses on decades except for the two 
last periods. Sweden had developed into a wealthy
welfare society with low income inequality,
particularly until the 1990s. A financial crisis struck Sweden in the early
1990s, which among other things increased unemployment and changed parts of
the welfare system towards being less generous. The start of the
last period (2008--2013) coincides with the international financial
crisis. During the last decades income inequality has increased considerably.

@@

# The effect of social class, 1801--1950, 1961--2013

@@

The cumulative
hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of
graphically illustrating the development over time.

@@

Notice that we have not analysed the 1950s due to the fact that we only
have data for some parishes at that time. Furthermore, the
results for women during the 1960s and 1970 are highly misleading for the reasons discussed above. The
changing economy also have consequences for the analysis, particularly when
it comes to the agricultural sector. While farmers were common occupations
up until the first half of the 20th century, these occupations has constantly
decreased in numbers thereafter. This means that the middle class has
a different character during the last periods.

The most striking result of the analysis is the different patterns for men
and women. The social pattern of mortality among women is quite much in
accordance with previous analyses from the present research group on other
regions in northern Sweden and for other periods [@edli11, @edbr12]. The
highest social class usually have comparatively low mortality, while
working class women have high, but almost constantly about the same level as the middle category. There is however no substantial change in the relative risks between classes. Only the group with unknown social class is distinguished by much higher mortality from the 1980s
onwards. This can however have to do with how we have defined social
position for women before and after 1950. After 1960 it is based
on their own occupation. Being without occupation probably include women that
are more frail when it comes to health or are vulnerable due to social characteristics.

A completely different pattern appears among men, being the most surprising finding in
our analyses. The groups we would expect to have the lowest mortality, if
we assume that access to vital resources and belonging to a high social class
determines survival, instead have by far the highest hazards of dying
during the nineteenth and much of the 20th century. Workers on the
other hand have comparatively low mortality until the 1960s, at the same
level as the middle class. The relative positions change radically towards
the end the 20th century. From the 1980s, there is a clear advantage when
it comes to survival for higher managers and professionals. The relative
differences are however more compressed than for the period until 1950. The
expected social gradient in mortality has developed. This is also what we
usually find in analyses of social health inequalities in present-day
Sweden as well as in most other countries. In the regions we are studying,
this is however a quite recent phenomenon. Corresponding to what we find
for women, those without known occupation have much higher mortality than
the others from the 1960s onwards. The assumption made for women of
unknown social class is probably even more relevant for men. They probably
represent groups in society that are particularly vulnerable.

@@

Concerning the social pattern, the pattern is less clear compared to what
we find for the younger age group (see Figure \@ref(fig:cumswold)). For
women the results are mixed. Once again, too much missing data on social
class force us to leave out the results for the 1960s and 1970s. We do not find the
same clear advantage for elite women during the period before 1950, but neither
do we find any large disadvantage for working class elderly women at that
time. In some periods, the social differences were small, particularly between the worker and the middle group, and the elite even had substantially higher mortality during the early 20th century. The
social differences continued to be small until the last decades of the
20th century. Only during the two last periods (after the year 2000) is it
possible to establish the expected social gradient in mortality.

@@

In this and the following section, the focus is on analyses of mortality from the two
dominant groups of death causes in the adult and elderly population. We start
with cardiovascular diseases. Many studies have shown that socio-economic
conditions are strongly associated with mortality in this disease
group. Belonging to working class, having less education and low income
increases mortality risks. We would therefore expect to find a clear
social pattern in the present analysis.

@@

As before, models are fitted separately for sex and each ten-year time period, from  1901 to
2013. We stratify on social class (**hisclass**) and include the covariates **urban** (TRUE or FALSE) and
**civst** (marital status). The cumulative
hazards at the end of the age period (65 if 40--64, 90 if 65--89) are the
main target in our analysis. These numbers are used in the process of
graphically illustrating the development over time.

@@

Figure \@ref(fig:41cumsw) shows mortality levels during different periods
from 1901 on-wards and the result of the model can be seen in Table \@ref(tab:41cumswtab). Women from the highest social class had consistently
lower mortality from cardiovascular diseases compared to the others. We
also observe increasing social differences from the first half of the
century to the last decades. A new pattern has developed from 1980
onwards. Now working class women have substantially higher cardiovascular
mortality compared to others. We can however not exclude that this is an
artefact of different diagnostics for deaths in different groups. Working
class women may have had less precise diagnoses in the early 20th century.

For men, there is a different pattern, a pattern however that to a large
extent is consistent with what we observed concerning all-cause
mortality. The elite group had much higher cardiovascular mortality up until
the middle of the 20th century, but instead have the lowest mortality from
1980 onwards. Working class men had a fairly low cardiovascular mortality,
but in the same way as among women, the levels were fairly similar to the
one of the middle category consisting of white collar, lower managers and farmers. In the last decades and similarly as for women, working class men have the highest
mortality from this cause of death group. A social
hierarchy in cardiovascular mortality developed both for men
and women.

@@

\section{Discussion}
